---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Johnson, Kevin"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```



-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",")


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

### Test Items starts from here - There are 6 sections ##########################

##### Section 1: (6 points) Summarizing the data.

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use table() to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}

#Present descriptive statistics
summary(mydata)

#Create a frequency table
table(mydata$CLASS, mydata$RINGS)
```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***The first variable shown is sex.  This is a categorical variable describing males, females and infants.  The first thing to observe is that there is a fair amount more males than both females and infants.  This is something to keep in mind going forward with the analysis because it could possible skew the data. The next two variables to observe are length and diameter.  What immediately sticks out for both of these are the mean and median are close to each other, hinting they are likely evenly distributed.  The same could be said for height.  Whole weight looks to be close to normally distributed.  However, one thing that stands out is the large range.  This could potentially point to outliers.  Shuck has similar characteristics to weight.  Rings are a rough estimation of age, hence the integers.  What is interesting about the statistics for rings are that mean is nearly a full integer above median.  Typically, one would expect a normal distribution for age.  However, this data is skewed to the right.  Class is another categorical variable that groups based off rings/age.  Volume exhibits similar characteristics to whole and shuck weight.  Ratio calculate shuck by volume, so it will show similar characteristics to those variables.***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
#Create a table of sex by class
sex_by_class <- table(mydata$SEX, mydata$CLASS)
addmargins(sex_by_class)

#Create a bar plot of sex by class
library(RColorBrewer)
barplot(sex_by_class, beside = , col = brewer.pal(3, "Greens"), border = "black",
        main = "Abalones Grouped by Sex", xlab = "Classes grouped by sex", ylab = "Count")

```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Class is a categorical variable that groups by age, with A1 being the youngest and A5 being the oldest.  With that in mind, it is intersting that not all of A1 consists of infant.  Looking at A2, one could presume that around this age is when most abalones begin to show their sex.  A3 stands out for the large amount of males and females.  Females are overwhelmingly represented in this category.  A4 and A5 appear to have similar proportions for each sex.***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}

#Select a simple random sample of 200 observations
set.seed(123)
mydata_ints <- c(1:nrow(mydata))
sample_rows <- sample(mydata_ints, 200)
work <- mydata[sample_rows,]

# Create a scatterplot matrix of variables 2-6
plot(work[, 2:6], pch = 16, , cex = 0.7, col=rgb(1, 3, 5, maxColorValue = 255))
```

-----

##### Section 2: (5 points) Summarizing the data using graphics.

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
plot(mydata$WHOLE, mydata$VOLUME, pch = 16, col=brewer.pal(5, "Spectral")[mydata$CLASS],
     main = "Comparison of Weight of Whole to Volume",
     xlab=" Whole Weight (grams)", ylab="Volume")
legend(x="bottomright", legend=unique(mydata$CLASS),
       col=brewer.pal(5, "Spectral"), pch=16)

```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}

plot(mydata$WHOLE, mydata$SHUCK, pch = 16, col=brewer.pal(5, "Spectral")[mydata$CLASS],
     main = "Comparison of Weight of Whole to Shucked Weight",
     xlab="Whole Weight (grams)", ylab="Shucked Weight (grams)")
abline(0, max(mydata$SHUCK/mydata$WHOLE), col=rgb(1, 3, 5, maxColorValue = 255))

max_ratio <- which.max(mydata$SHUCK/mydata$WHOLE)
text(mydata[max_ratio,]$WHOLE, mydata[max_ratio,]$SHUCK*1.15, 
  "Max Ratio", cex = .8)
legend(x="bottomright", legend=unique(mydata$CLASS),
       col=brewer.pal(5, "Spectral"), pch=16)

```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Both plots in A and B have a similar shape to each other.  Both also exhibit similar positive correlations.  There is a very clear tight correlation at lower weights.  However, the heavier, the abalone gets, the more dispersed they get.Looking at different age classes, it is not surprising that the younger abalones will have lower weights and volumes.  Meanwhile, the adult adalones all appear to be clustered between 100 and 200 grams.  That being said, it is apparent there are quite a few outliers in the data.  However, that may be expected given the right skew.***

-----

##### Section 3: (8 points) Getting insights about the data using graphs.

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}

# Create a histogram of ratio differentiated by sex
mydata_hist <- ggplot(data = mydata, aes(x = RATIO, fill=SEX)) +
  geom_histogram(binwidth=.02) +
  facet_grid(. ~ SEX) +
  scale_fill_brewer(palette="Set2") +
  labs(x = "Ratio", y="Count", color = "Sex")

# Create a boxplot of ratio differentiated by sex
mydata_box <- ggplot(data = mydata, aes(y = RATIO, color=SEX)) +
  geom_boxplot(width=1) +
  xlim(-0.5, 0.5) +
  facet_grid(. ~ SEX) +
  scale_color_brewer(palette="Set2") +
  labs(x = "", y="Ratio", color = "Sex")

# Create a Q-Q plot of ratio differentiated by sex
mydata_qq <- ggplot(data = mydata, aes(sample = RATIO, color=SEX)) +
  geom_qq() +
  geom_qq_line() +
  facet_grid(. ~ SEX) +
  scale_color_brewer(palette="Set2") +
  labs(x = "Theoretical", y="Sample Ratio", color = "Sex")
grid.arrange(mydata_hist, mydata_box, mydata_qq)
```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***The first thing that sticks out is the slightly right skew of the three charts.  The male histogram is the closest to a bell curve shape, however all three categories are lsightly leaning to the right.  The boxplots show to be close to the mean.  However, there are still some outliers, with females being the most pronounced.  The Q-Q plot shows most of the data sticks close to the line.  However, there is a right tail to the data that disperse from the Q-Q line.***

(3)(b) (2 points) Use the boxplots to identify RATIO outliers (mild and extreme both) for each sex. Present the abalones with these outlying RATIO values along with their associated variables in "mydata" (Hint:  display the observations by passing a data frame to the kable() function).  

```{r Part_3b}

#Outliers function
mydata_outliers <- function(sex){
  return(boxplot.stats(x = mydata$RATIO[mydata$SEX == sex])[["out"]])
}

#Extreme outliers function
mydata_extreme <- function(outliers, sex){
  return(outliers > (quantile(mydata$RATIO[mydata$SEX == sex], .75) + 3.0*IQR(mydata$RATIO[mydata$SEX == sex])))
}

#Find ratio outliers for each sex
ratio_outliers_male <- mydata_outliers("M")
ratio_outliers_female <- mydata_outliers("F")
ratio_outliers_infant <- mydata_outliers("I")

#Output outliers
library(kableExtra)
ratio_outliers_pos <- match(c(ratio_outliers_male, ratio_outliers_female, ratio_outliers_infant), mydata$RATIO)
kable(mydata[ratio_outliers_pos, ], format = "html", caption = "Ratio Outliers by Sex") %>% kable_styling()

#Find extreme outliers for each sex
ratio_extreme_male <- mydata_extreme(ratio_outliers_male, "M")
ratio_extreme_female <- mydata_extreme(ratio_outliers_female, "F")
ratio_extreme_infant <- mydata_extreme(ratio_outliers_infant, "I")

#Output extreme outliers
ratio_outliers_extreme <- c(ratio_outliers_male[ratio_extreme_male],ratio_outliers_female[ratio_extreme_female],ratio_outliers_infant[ratio_extreme_infant])
ratio_extreme_pos <- match(ratio_outliers_extreme, mydata$RATIO)
kable(mydata[ratio_extreme_pos, ], format = "html", caption = "Ratio Extreme Outliers by Sex") %>% kable_styling()

```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***There are 19 total outliers.  This may seem like a lot at first glance but it may be expected due to the right skew of the data. Of the 19 outliers, there are 5 males, 6 females, and 8 infants.  The larger amount of infants could be explained by the smaller volume and weights of most infants.  This can be said for the two extreme outliers as well, seeing as the female is in Class A2 and the infant is Class A1.  ***

-----

##### Section 4: (8 points) Getting insights about possible predictors.

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
#Create a boxplot of volume differentiated by class
mydata_volumebox <- ggplot(data = mydata, aes(y = VOLUME, x = CLASS, color=CLASS)) +
  geom_boxplot() +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Class", y="Volume", color = "Class")

#Create a boxplot for whole differentiated by class
mydata_wholebox <- ggplot(data = mydata, aes(y = WHOLE, x = CLASS, color=CLASS)) +
  geom_boxplot() +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Class", y="Whole Weight (grams)", color = "Class")

#Create a scatterplot for volume and split by rings
mydata_volumescatter <- ggplot(data = mydata, aes(y = VOLUME, x = RINGS, color=CLASS)) +
  geom_point() +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Rings", y="Volume", color = "Class")

#Create a scatterplot for whole and split by rings
mydata_wholescatter <- ggplot(data = mydata, aes(y = WHOLE, x = RINGS, color=CLASS)) +
  geom_point() +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Rings", y="Whole Weight (grams)", color = "Class")
grid.arrange(mydata_volumebox, mydata_wholebox, mydata_volumescatter, mydata_wholescatter, ncol=2)


```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***There is initially a strong correlation between rings with volume and weight.  However, the higher the class goes, this begins to fade away.  There is much more variability within the data for Classes A3, A4, and A5.  Much of the weight and volumes in those three classes tend to be pretty close as well.  Therefore, I do not think these variables would perform well as predictors of age.  That being said, however, they could potentially be good predictors for determining infants verse adults.***

-----

##### Section 5: (12 points) Getting insights regarding different groups in the data.

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
#Compute the mean values of volume, shuck, and ratio
mydata_means <- aggregate(x = mydata[,c("VOLUME","SHUCK","RATIO")], by = list(mydata$SEX, mydata$CLASS), FUN = "mean")

#Create volume matrix
volume_matrix <- matrix(mydata_means$VOLUME, nrow = 3, ncol = 5, dimnames = list(unique(mydata_means$Group.1), unique(mydata_means$Group.2)))
kable(volume_matrix, format = "html", caption = "Mean Volume by Sex and Class") %>% kable_styling()

#Create shuck matrix
shuck_matrix <- matrix(mydata_means$SHUCK, nrow = 3, ncol = 5, dimnames = list(unique(mydata_means$Group.1), unique(mydata_means$Group.2)))
kable(volume_matrix, format = "html", caption = "Mean Shuck Weight by Sex and Class") %>% kable_styling()

#Create ratio matrix
ratio_matrix <- matrix(mydata_means$RATIO, nrow = 3, ncol = 5, dimnames = list(unique(mydata_means$Group.1), unique(mydata_means$Group.2)))
kable(ratio_matrix, format = "html", caption = "Mean Shuck Weight to Volume Ratio by Sex and Class") %>% kable_styling()

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}

# Create ratio versus class line graph
ggplot(data = mydata_means, aes(y = RATIO, x = mydata_means$Group.2, group = Group.1, color=Group.1)) + 
  geom_line(size = 0.8) +
  geom_point(size = 2) +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Class", y="Ratio of Shuck Weight to Volume", color = "Sex", title = "Mean Ratio of Shuck Weight to Volume by Class")

#Create volume versus class line graph
ggplot(data = mydata_means, aes(y = VOLUME, x = mydata_means$Group.2, group = Group.1, color=Group.1)) +
  geom_line(size = 0.8) +
  geom_point(size = 2) +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Class", y="Volume", color = "Sex", title = "Mean Volume by Class")

#Create shuck versus class line graph
ggplot(data = mydata_means, aes(y = SHUCK, x = mydata_means$Group.2, group = Group.1, color=Group.1)) +
  geom_line(size = 0.8) +
  geom_point(size = 2) +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Class", y="Shuck Weight (grams)", color = "Sex", title = "Mean Shuck Weight by Class")
```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***I find the clear negative correlation between ratio of shuck weight to volume with age to be very interesting.  It appears that as abalones age they gain volume but not weight. There could be a possible future experiment to explain why for this occurring.  You can also see on the second graph that abalones gain volume as they age.  However, that begins to slow down between Class A3 and A4 until remaining about the same in A5.   ***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}

#Subset the data
mydata_infant <- subset(mydata, RINGS < 16 & SEX == 'I')
mydata_adult <- subset(mydata, RINGS < 16 & (SEX == 'M' | SEX == 'F'))
par(mfrow = c(2, 2))

#Create infant boxplot for volume
boxplot(VOLUME ~ RINGS, data = mydata_infant, col = rgb(1, 3, 5, maxColorValue = 255),
        ylim = c(0, 1100), main = "Volume of Infants by Rings", ylab = "Volume")

#Create adult boxplot for volume
boxplot(VOLUME ~ RINGS, data = mydata_adult, col = rgb(1, 3, 5, maxColorValue = 255),
        ylim = c(0, 1100),main = "Volume of Adults by Rings", ylab = "Volume")

#Create infant boxplot for whole weight
boxplot(WHOLE ~ RINGS, data = mydata_infant, col = rgb(1, 3, 5, maxColorValue = 255),
        ylim = c(0, 400),main = "Whole Weight of Infants by Rings", ylab = "Whole Weight")

#Create adult boxplot for whole weight
boxplot(WHOLE ~ RINGS, data = mydata_adult, col = rgb(1, 3, 5, maxColorValue = 255),
        ylim = c(0, 400), main = "Whole Weight of Adults by Rings", ylab = "Whole Weight")
```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***One thing that stands out in the adult plots is the larger amount of variability for weights compared to rings.  Not only are the ranges wider, the outliers are larger as well.  You can also tell there are not many adult abalones with a low amount of rings.  Somewhere between 8 to 10 are when most abalones jump from infant to adult.  Looking at the adult plots, you can see a positive correlation with rings to volume and weight.  However, this tends to level off with age, and at about 9 or 10 this correlation ceases to exist for most abalones.***

-----

##### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).

**Conclusions**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***First of all, the skew of the data could be a red flag for why the original study failed.  It is possible there was not an effective sample taken for analysis.  Males also appeared to be overrepresented in the analysis.  There seemed to be some classification issues within the data as well.  For example, though Classes A4 and A5 are meant to be the oldest of the populationm there were still a fair amount of infants showing.  Similarly, there were adults in Class A1, and a lot in Class A2 as well.  Correct classification is important for this type of study, so the fact there are so many concerns regarding this leads to a lot of questions about the accuracy of the analysis.***

**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***I typically would expect something close to a normal distribution in order to perform an accurate analysis.  The mean and median should be close to each other in the summary startics.  In the histogram, I would expect a shape similar to the bell curve.  Categorical showing strong representation for each group would be important as well.  The range and variation of the data may be important as well.  However, this may change depending on the size of the population.  Typically, I would hope for a population within the data for a stronger analysis.***

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***The most important questions when analyzing observational studies would be the method of data collection and the population being observed.  Both of these could have a strong affect on the data.  The researcher could have bias as well.  These questions, specifically, are why I think domain knowledge is important for data scientists.  The better the analyst understands what is being observed, the better they can perform experiments on the data. Similarly, if a data scientist has a strong domain knowledge, they will be better equipped to spot flaws in the data before it it too late.***
